---
title: "Methodology and Current Implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_method_overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

This vignette describes the methodology implemented by the current
`causalBKMR` package. The deployed API lives in the active `R/` files:

- `prepare_gbkmr_data()` converts user matrices into the wide g-BKMR
  analysis data.
- `detect_variable_patterns()` audits the resulting columns.
- `gbkmr_run()` runs the standard analysis.
- `run_gbkmr_panel()` is a lower-level exported function for advanced
  users who already understand the package's internal wide-data
  conventions.

Archived code under `R/old/` is not used in the vignettes or the current
recommended workflow.

## Problem Setup

We observe `n` subjects over `T` time points. At each time point, each
subject has:

- `p` mixture exposure components.
- `Ldim` time-varying covariates.
- optional baseline covariates.
- an outcome `Y` measured after the exposure and covariate history.

The user supplies three objects:

| Object | Shape | Meaning |
|--------|-------|---------|
| `Y` | length `n` | Continuous or binary outcome |
| `Z` | `n x (p * T)` | Mixture exposures ordered by time |
| `X` | `n x (Ldim * T + B)` | Time-varying covariates first, baseline covariates last |

`prepare_gbkmr_data()` converts those objects into a wide data frame with
columns such as `logM1_0`, `logM2_0`, `waist_0`, `logM1_1`, and `Y`.

```{r prepare-example, purl=FALSE}
prepared <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = 3,
  mixture_components = 2,
  td_covariates = 1,
  baseline_covariates = 1,
  td_covariate_names = "waist"
)

detect_variable_patterns(prepared, T = 3)
```

## Causal Estimand

The target estimand is the average causal effect comparing two full
mixture trajectories:

$$
ACE = E[Y(a^*)] - E[Y(a)].
$$

By default, `gbkmr_run()` sets `a` to the 25th percentile and `a*` to the
75th percentile for every mixture component at every time point:

```{r default-contrast, purl=FALSE}
fit <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_probs = c(0.25, 0.75)
)
```

Users can supply explicit trajectories with `a_vals` and `astar_vals`.
Those vectors must be named with the mixture column names in the
prepared data.

```{r explicit-contrast, purl=FALSE}
exposure_names <- grep("^logM\\d+_\\d+$", names(prepared), value = TRUE)

a_vals <- setNames(rep(0.5, length(exposure_names)), exposure_names)
astar_vals <- setNames(rep(1.5, length(exposure_names)), exposure_names)

fit_custom <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_vals = a_vals,
  astar_vals = astar_vals
)
```

## Model Sequence

The current `gbkmr_run()` implementation follows this sequence:

1. Detect mixture, time-varying, baseline, and outcome columns.
2. Fit mediator models for time-varying covariates after baseline.
3. Fit the outcome model.
4. Simulate counterfactual covariate histories under `a` and `a*`.
5. Predict counterfactual outcomes.
6. Return posterior summaries of `Y(a*) - Y(a)`.

For continuous outcomes, the result is on the outcome scale. For binary
outcomes, the result is a risk difference.

## BKMR Engines

`gbkmr_run()` can use two engines:

| Engine | Function | Use case |
|--------|----------|----------|
| `bkmr` | `bkmr::kmbayes()` | Default engine; supports continuous and binary outcomes |
| `fastbkmr` | `fbkmr::skmbayes()` | Optional large-sample engine for continuous outcomes |

With `engine = "auto"`, the package uses `fastbkmr` only for large
continuous-outcome analyses when `fbkmr` is installed. Otherwise it uses
standard `bkmr`.

Version 1 support note:

| Outcome | Time-dependent confounder `L` | `engine = "bkmr"` | `engine = "fastbkmr"` |
|---------|-------------------------------|-------------------|-----------------------|
| continuous | all continuous | Supported: Gaussian models throughout | Supported: Gaussian models throughout |
| binary | all continuous | Supported: probit outcome model | Not supported; hard error |
| continuous | any binary | Runs with warning: binary `L` is fit and sampled as Gaussian | Runs with warning: binary `L` is fit and sampled as Gaussian |
| binary | any binary | Runs with warning: probit outcome plus Gaussian `L` models | Not supported; hard error |

In this table, binary means a 0/1 outcome or a 0/1 time-dependent
confounder. Mixture exposures can remain continuous. The current public
`fbkmr::skmbayes()` path is Gaussian-only; generalized/logistic fastBKMR
work is not part of the installed `fbkmr` package used here.

## Assumptions

The causal interpretation depends on standard longitudinal g-formula
conditions:

- consistency of the observed outcome under the observed exposure
  history;
- sequential exchangeability after conditioning on measured history;
- positivity for the intervention trajectories;
- correct enough model specification for the fitted BKMR mediator and
  outcome models.

The flexible kernel helps with nonlinear and interactive exposure
response, but it does not remove the need for measured confounding
control and realistic intervention values.

## Practical Limits

- Baseline covariates enter as linear terms rather than kernel
  modifiers.
- Time-varying covariates are modeled as Gaussian mediator responses in
  the current implementation.
- Binary outcomes are supported only by the standard `bkmr` engine.
- Explicit `a_vals` and `astar_vals` should stay within observed support
  unless extrapolation is scientifically justified.

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Binary outcomes](gbkmr_example_binary.html)
- [Large samples with fastBKMR](gbkmr_example_fastbkmr.html)
- [Diagnostics and troubleshooting](gbkmr_diagnostics.html)
