---
title: "Binary Outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_example_binary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

This vignette uses the current `causalBKMR` API for a binary outcome.
The workflow is the same as for a continuous outcome:

1. Build `Y`, `Z`, and `X`.
2. Convert them with `prepare_gbkmr_data()`.
3. Run the analysis with `gbkmr_run(outcome_type = "binary")`.

Archived functions under `R/old/` are not used by this workflow.

## Data Layout

`gbkmr_run()` expects the wide data produced by `prepare_gbkmr_data()`.
For `T = 3`, `p = 2`, one time-varying covariate, and one baseline
covariate, the raw matrices are:

- `Y`: length `n` binary vector with values 0 or 1.
- `Z`: `n x 6` matrix ordered as exposure 1 to 2 at time 0, then time 1,
  then time 2.
- `X`: `n x 4` matrix with the three time-varying covariate columns first
  and the baseline covariate column last.

```{r simulate-binary, purl=FALSE}
library(causalBKMR)

set.seed(7)
n <- 300
T_points <- 3
p <- 2
Ldim <- 1
B <- 1

sex <- rbinom(n, 1, 0.5)
waist_0 <- rnorm(n, 85, 12)

Z <- matrix(rlnorm(n * p * T_points, meanlog = 1, sdlog = 0.5),
            nrow = n)

waist_t0 <- waist_0 + 0.10 * Z[, 1] + rnorm(n, 0, 2)
waist_t1 <- waist_t0 + 0.10 * Z[, 3] + rnorm(n, 0, 2)
waist_t2 <- waist_t1 + 0.10 * Z[, 5] + rnorm(n, 0, 2)

X <- cbind(waist_t0, waist_t1, waist_t2, sex)

linpred <- -3 +
  0.40 * sex +
  0.02 * waist_t2 +
  0.20 * log1p(Z[, 5]) +
  0.20 * log1p(Z[, 6])

Y <- rbinom(n, 1, plogis(linpred))
```

## Prepare the Wide g-BKMR Data

```{r prepare-binary, purl=FALSE}
prepared <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = T_points,
  mixture_components = p,
  td_covariates = Ldim,
  baseline_covariates = B,
  td_covariate_names = "waist"
)

detect_variable_patterns(prepared, T = T_points)
```

## Fit a Binary Outcome Model

Binary outcomes are supported with the standard `bkmr` engine. The
outcome model uses `bkmr::kmbayes(..., family = "binomial")`, and
`causalBKMR` returns counterfactual means on the probability scale.

```{r run-binary, purl=FALSE}
fit_binary <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "binary",
  time_points = T_points,
  engine = "bkmr",
  n = nrow(prepared),
  iter = 15000,
  K = 1000,
  n_knots = 50
)
```

The causal effect is a risk difference:

```{r inspect-binary, purl=FALSE}
fit_binary$causal_effect
fit_binary$counterfactual_means
fit_binary$diagnostics
```

If `fit_binary$causal_effect$estimate` is `0.04`, the high exposure
trajectory has an estimated four percentage point higher risk than the
low exposure trajectory.

## Engine Restrictions

`fastbkmr` does not currently support binary outcomes because the
current public `fbkmr::skmbayes()` path is Gaussian-only.

```{r binary-fastbkmr, purl=FALSE}
gbkmr_run(
  data = prepared,
  outcome_type = "binary",
  time_points = T_points,
  engine = "fastbkmr"
)
```

That call intentionally stops with a clear error. Use `engine = "bkmr"`
for binary outcomes.

## Time-Varying Covariates

Binary time-varying covariates can be included in `X`, but the current
mediator models treat time-varying covariates as Gaussian responses.
For binary intermediate covariates, interpret the mediator part of the
model as an approximation and check diagnostics carefully.

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Diagnostics and troubleshooting](gbkmr_diagnostics.html)
