---
title: "Custom Intervention Contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_example_contrasts}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

The current `gbkmr_run()` interface supports custom intervention
contrasts directly. You do not need to call any function from `R/old/`,
and you usually do not need to call the lower-level `run_gbkmr_panel()`.

## Default Contrast

By default, `gbkmr_run()` compares all mixture columns at the 25th
percentile with all mixture columns at the 75th percentile:

```{r default, purl=FALSE}
fit_default <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_probs = c(0.25, 0.75),
  iter = 15000,
  K = 1000
)
```

This means:

```text
low scenario:  logM1_0, logM2_0, ..., logM2_2 at their 25th percentiles
high scenario: logM1_0, logM2_0, ..., logM2_2 at their 75th percentiles
```

## Change Quantiles

Use `a_probs` for a different global quantile shift:

```{r quantiles, purl=FALSE}
fit_10_90 <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_probs = c(0.10, 0.90),
  iter = 15000,
  K = 1000
)
```

The two probabilities must be length two, with the low scenario first
and the high scenario second.

## Use Fixed Exposure Values

Use `a_vals` and `astar_vals` when the scientific question is about
specific exposure levels. These vectors must be named with the mixture
columns created by `prepare_gbkmr_data()`.

```{r fixed-values, purl=FALSE}
exposure_names <- grep("^logM[0-9]+_[0-9]+$", names(prepared), value = TRUE)

# prepare_gbkmr_data(log_transform_mixtures = TRUE) stores log concentrations.
a_background <- setNames(rep(log(1), length(exposure_names)), exposure_names)
a_high <- setNames(rep(log(5), length(exposure_names)), exposure_names)

fit_fixed <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_vals = a_background,
  astar_vals = a_high,
  iter = 15000,
  K = 1000
)
```

If `log_transform_mixtures = FALSE`, pass values on the original scale.
If it is `TRUE`, pass values on the log scale because the fitted data
contain log-transformed mixture columns.

## Change One Component

The names let you define component-specific and time-specific scenarios.
This example increases only component 2 at every visit:

```{r component-specific, purl=FALSE}
a_component <- setNames(rep(log(1), length(exposure_names)), exposure_names)
astar_component <- a_component
astar_component[c("logM2_0", "logM2_1", "logM2_2")] <- log(5)

fit_component <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_vals = a_component,
  astar_vals = astar_component,
  iter = 15000,
  K = 1000
)
```

## What Not To Include

Only mixture columns belong in `a_vals` and `astar_vals`. Do not include
baseline covariates or time-varying confounders:

```{r wrong, purl=FALSE}
# Wrong: baseline_1 is a confounder, not a mixture exposure.
a_wrong <- c(logM1_0 = 0, logM2_0 = 0, baseline_1 = 80)
```

Changing confounders inside the intervention vector would change the
estimand. In g-computation, time-varying confounders are sampled forward
under each mixture intervention.

## Compare Multiple Contrasts

```{r compare, purl=FALSE}
contrast_grid <- list(
  q25_q75 = c(0.25, 0.75),
  q10_q90 = c(0.10, 0.90),
  q05_q95 = c(0.05, 0.95)
)

fits <- lapply(contrast_grid, function(prob) {
  gbkmr_run(
    data = prepared,
    time_points = 3,
    a_probs = prob,
    iter = 15000,
    K = 1000
  )
})

do.call(rbind, lapply(names(fits), function(name) {
  ce <- fits[[name]]$causal_effect
  data.frame(
    contrast = name,
    estimate = ce$estimate,
    lower = ce$lower,
    upper = ce$upper
  )
}))
```

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Methodology overview](gbkmr_method_overview.html)
- [Diagnostics and troubleshooting](gbkmr_diagnostics.html)
