---
title: "Methodology and Current Implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_method_overview}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

This vignette describes the methodology implemented by the current
`causalBKMR` package. The deployed API lives in the active `R/` files:

- `prepare_gbkmr_data()` converts user matrices into the wide g-BKMR
  analysis data.
- `detect_variable_patterns()` audits the resulting columns.
- `gbkmr_run()` runs the standard analysis.
- `run_gbkmr_panel()` is a lower-level exported function for advanced
  users who already understand the package's internal wide-data
  conventions.

Archived code under `R/old/` is not used in the vignettes or the current
recommended workflow.

## Problem Setup

We observe `n` subjects over `T` time points. At each time point, each
subject has:

- `p` mixture exposure components.
- `Ldim` time-varying covariates.
- optional baseline covariates.
- an outcome `Y` measured after the exposure and covariate history.

The user supplies three objects:

| Object | Shape | Meaning |
|--------|-------|---------|
| `Y` | length `n` | Continuous or binary outcome |
| `Z` | `n x (p * T)` | Mixture exposures ordered by time |
| `X` | `n x (Ldim * (T - 1) + B)` | Post-baseline time-varying covariates first, baseline covariates last |

`prepare_gbkmr_data()` converts those objects into a wide data frame with
columns such as `baseline_1`, `logM1_0`, `logM2_0`, `waist_1`, and `Y`.

```{r prepare-example, purl=FALSE}
prepared <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = 3,
  mixture_components = 2,
  td_covariates = 1,
  baseline_covariates = 1,
  td_covariate_names = "waist"
)

detect_variable_patterns(prepared, T = 3)
```

## Causal Estimand

The target estimand is the average causal effect comparing two full
mixture trajectories:

$$
ACE = E[Y(a^*)] - E[Y(a)].
$$

By default, `gbkmr_run()` sets `a` to the 25th percentile and `a*` to the
75th percentile for every mixture component at every time point:

```{r default-contrast, purl=FALSE}
fit <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_probs = c(0.25, 0.75)
)
```

Users can supply explicit trajectories with `a_vals` and `astar_vals`.
Those vectors must be named with the mixture column names in the
prepared data.

```{r explicit-contrast, purl=FALSE}
exposure_names <- grep("^logM[0-9]+_[0-9]+$", names(prepared), value = TRUE)

a_vals <- setNames(rep(0.5, length(exposure_names)), exposure_names)
astar_vals <- setNames(rep(1.5, length(exposure_names)), exposure_names)

fit_custom <- gbkmr_run(
  data = prepared,
  time_points = 3,
  a_vals = a_vals,
  astar_vals = astar_vals
)
```

## Model Sequence

The current `gbkmr_run()` implementation follows this sequence:

1. Detect mixture, time-varying, baseline, and outcome columns.
2. Fit models for post-baseline time-varying confounders.
3. Fit the outcome model.
4. Simulate counterfactual covariate histories under `a` and `a*`.
5. Predict counterfactual outcomes.
6. Return posterior summaries of `Y(a*) - Y(a)`.

For continuous outcomes, the result is on the outcome scale. For binary
outcomes, the result is a risk difference.

## BKMR Engines

`gbkmr_run()` can use two engines:

| Engine | Function | Use case |
|--------|----------|----------|
| `bkmr` | `bkmr::kmbayes()` | Default engine; supports continuous and binary outcomes, including binary time-dependent confounders |
| `fastbkmr` | `fbkmr::skmbayes()` | Optional large-sample engine for all-Gaussian analyses only |

With `engine = "auto"`, the package uses `fastbkmr` only for large
analyses with a continuous outcome and all continuous time-dependent
confounders when `fbkmr` is installed. Otherwise it uses standard `bkmr`.

Version 1 support note:

| Outcome | Time-dependent confounder `L` | `engine = "bkmr"` | `engine = "fastbkmr"` |
|---------|-------------------------------|-------------------|-----------------------|
| continuous | all continuous | Supported: Gaussian outcome + Gaussian `L` models | Supported: Gaussian fastBKMR throughout |
| binary | all continuous | Supported: probit outcome + Gaussian `L` models | Not supported; hard error |
| continuous | any binary | Supported: Gaussian outcome + probit models for binary time-varying confounders with Bernoulli Monte Carlo sampling | Not supported; hard error |
| binary | any binary | Supported: probit outcome + probit models for binary time-varying confounders with Bernoulli Monte Carlo sampling | Not supported; hard error |

In this table, binary means a 0/1 outcome or a 0/1 time-dependent
confounder. Mixture exposures can remain continuous. The current public
`fbkmr::skmbayes()` path is Gaussian-only; generalized/logistic fastBKMR
work is not part of the installed `fbkmr` package used here.

## Assumptions

The causal interpretation depends on standard longitudinal g-formula
conditions:

- consistency of the observed outcome under the observed exposure
  history;
- sequential exchangeability after conditioning on measured history;
- positivity for the intervention trajectories;
- correct enough model specification for the fitted BKMR models for
  time-varying confounders and outcomes.

The flexible kernel helps with nonlinear and interactive exposure
response, but it does not remove the need for measured confounding
control and realistic intervention values.

## Practical Limits

- Baseline covariates enter as linear terms rather than kernel
  modifiers.
- Binary time-dependent covariates use probit BKMR models and
  Bernoulli Monte Carlo sampling under the standard `bkmr` engine.
- Binary outcomes and binary time-dependent covariates are supported only
  by the standard `bkmr` engine.
- Explicit `a_vals` and `astar_vals` should stay within observed support
  unless extrapolation is scientifically justified.

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Binary outcomes](gbkmr_example_binary.html)
- [Large samples with fastBKMR](gbkmr_example_fastbkmr.html)
- [Diagnostics and troubleshooting](gbkmr_diagnostics.html)
