---
title: "Diagnostics and Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_diagnostics}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

This vignette covers diagnostics for the current `causalBKMR` workflow.
The relevant user-facing functions are `prepare_gbkmr_data()`,
`detect_variable_patterns()`, and `gbkmr_run()`. Archived code under
`R/old/` is not part of this workflow.

## Check the Prepared Data

Before running MCMC, inspect the columns that `gbkmr_run()` will use.

```{r check-patterns, purl=FALSE}
patterns <- detect_variable_patterns(prepared, T = 3)
patterns

mixture_cols <- grep("^logM[0-9]+_[0-9]+$", names(prepared), value = TRUE)
mixture_cols
```

Confirm that:

- `patterns$p` is the expected number of mixture components per time.
- `mixture_cols` contains the expected `logM*_t` columns.
- `patterns$Ldim` is the expected number of time-varying covariates per time.
- `patterns$td_covariate_names` contains the expected covariate names.
- `patterns$td_vars_by_time` maps the follow-up covariate columns correctly.
- `Y` is present and points to the intended outcome column.

If the detected columns are wrong, return to the `Y`, `Z`, and `X`
matrices passed into `prepare_gbkmr_data()`. The most common issue is an
incorrect ordering of `X`: all time-varying covariates must come before
baseline covariates.

## Use the Input Audit

`gbkmr_run(verbose = TRUE)` prints an audit before fitting models.

```{r audit, purl=FALSE}
fit <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  engine = "auto",
  iter = 15000,
  K = 1000,
  verbose = TRUE
)
```

Read this audit before starting a long run. Check the outcome type,
number of time points, number of mixture components, selected engine,
sample size, and intervention contrast.

## Inspect Stored Diagnostics

The result object stores convergence diagnostics for each fitted BKMR
model.

```{r stored-diagnostics, purl=FALSE}
fit$diagnostics
```

Each entry corresponds to one model. `outcome_Y` is the final outcome
BKMR model. Entries such as `confounder_t1_1` are time-dependent
confounder models; the name means the first confounder model at visit
`t = 1`.

The current diagnostics are computed from the posterior draws of the
linear covariate coefficients (`beta`) in each BKMR fit:

- `ess`: minimum effective sample size across the beta MCMC chains. ESS
  is the approximate number of independent posterior draws after
  accounting for autocorrelation. Larger values are better; values below
  100 are flagged as `low_ess`.
- `geweke_max_abs`: maximum absolute Geweke z-score across beta chains.
  This compares early and late parts of a chain. Values above 2.5 are
  flagged as `geweke_nonstationary`.
- `warning_flags`: a character vector of flags. An empty vector means no
  issue was found by these quick checks.

These are quick screening diagnostics, not a full MCMC review. They
should be used together with trace plots and posterior summary plots.

If warnings appear, rerun with more iterations and a later posterior
selection window:

```{r rerun-longer, purl=FALSE}
fit_long <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  iter = 60000,
  sel = seq(floor(60000 * 0.8), 60000, by = 25),
  K = 1000
)
```

The `sel` argument must contain MCMC iteration indices no larger than
`iter`. Use later iterations when early burn-in appears unstable.

## Inspect Raw Fits

`gbkmr_run()` returns lower-level model objects in `raw_results`.

```{r raw-results, purl=FALSE}
raw <- fit$raw_results

names(raw)
raw$fit_y
raw$meta
```

For the standard BKMR engine, `raw$fit_y` is a `bkmrfit` object. For
fastBKMR, `raw$fit_y` is a list of subset fits.

## Trace Plots

Use `bkmr::TracePlot()` for standard BKMR fits.

```{r trace-standard, purl=FALSE}
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "beta")
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "sigsq.eps")
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "r")
```

For fastBKMR, inspect subset fits:

```{r trace-fast, purl=FALSE}
fit_list <- fit$raw_results$fit_y

bkmr::TracePlot(fit = fit_list[[1]], par = "beta")

for (i in seq_along(fit_list)) {
  bkmr::TracePlot(fit = fit_list[[i]], par = "beta")
}
```

Long trends, abrupt jumps, or chains that remain stuck for long periods
are signs that the run needs more iterations or a different `sel`
window.

## BKMR Summary Plots

`gbkmr_run()` stores the fitted BKMR objects under `fit$raw_results`.
This means the usual `bkmr` summary functions can be used directly for
the outcome model and for each time-dependent confounder model.

The examples below use the standard BKMR engine. If `engine = "fastbkmr"`,
the stored fit is a list of subset fits; inspect one subset at a time or
loop over subsets.

```{r bkmr-plot-helpers, purl=FALSE}
first_bkmrfit <- function(x) {
  if (is.list(x) && !inherits(x, "bkmrfit")) x[[1]] else x
}

fit_z_names <- function(fit_object, fallback_prefix = "Z") {
  out <- colnames(fit_object$Z)
  if (is.null(out)) out <- paste0(fallback_prefix, seq_len(ncol(fit_object$Z)))
  out
}

sel <- fit$raw_results$meta$sel
```

For the outcome model, `raw_results$fit_y` is the BKMR fit for `Y`. Its
kernel predictors include all mixture exposure columns and the modeled
time-dependent confounder history. Baseline covariates are in the linear
`X` part of the BKMR model, not in the kernel `h()`.

```{r outcome-summary-plots, purl=FALSE}
outcome_fit <- first_bkmrfit(fit$raw_results$fit_y)

outcome_z_names <- c(
  fit$raw_results$meta$exposure_names,
  fit$raw_results$meta$confounder_names
)
if (length(outcome_z_names) != ncol(outcome_fit$Z)) {
  outcome_z_names <- fit_z_names(outcome_fit)
}

pips_y <- bkmr::ExtractPIPs(outcome_fit, sel = sel, z.names = outcome_z_names)
barplot(setNames(pips_y$PIP, pips_y$variable),
        las = 2, ylab = "Posterior inclusion probability",
        main = "Outcome model PIPs")

pr_y <- bkmr::PredictorResponseUnivar(
  fit = outcome_fit,
  Z = outcome_fit$Z,
  X = outcome_fit$X,
  sel = sel,
  z.names = outcome_z_names
)

one_var <- as.character(pr_y$variable[1])
pr_one <- pr_y[pr_y$variable == one_var, ]
plot(pr_one$z, pr_one$est, type = "l",
     xlab = one_var, ylab = "Estimated h contribution",
     main = paste("Outcome model:", one_var))
lines(pr_one$z, pr_one$est + 1.96 * pr_one$se, lty = 2)
lines(pr_one$z, pr_one$est - 1.96 * pr_one$se, lty = 2)

overall_y <- bkmr::OverallRiskSummaries(
  fit = outcome_fit,
  Z = outcome_fit$Z,
  X = outcome_fit$X,
  sel = sel,
  qs = seq(0.25, 0.75, by = 0.05)
)
plot(overall_y$quantile, overall_y$est, type = "b",
     xlab = "Joint mixture quantile",
     ylab = "Difference in h from median mixture",
     main = "Outcome model overall mixture summary")
lines(overall_y$quantile, overall_y$est + 1.96 * overall_y$sd, lty = 2)
lines(overall_y$quantile, overall_y$est - 1.96 * overall_y$sd, lty = 2)
```

For time-dependent confounder models, `raw_results$fit_confounders[[t]][[l]]`
is the BKMR model for the `l`th time-dependent confounder at follow-up
visit `t`. Its kernel predictors are the exposure and confounder history
available before that confounder is generated.

```{r confounder-summary-plots, purl=FALSE}
confounder_fit <- first_bkmrfit(fit$raw_results$fit_confounders[[1]][[1]])
confounder_z_names <- fit_z_names(confounder_fit)

pips_l <- bkmr::ExtractPIPs(
  confounder_fit,
  sel = sel,
  z.names = confounder_z_names
)
barplot(setNames(pips_l$PIP, pips_l$variable),
        las = 2, ylab = "Posterior inclusion probability",
        main = "Time-dependent confounder model PIPs")

pr_l <- bkmr::PredictorResponseUnivar(
  fit = confounder_fit,
  Z = confounder_fit$Z,
  X = confounder_fit$X,
  sel = sel,
  z.names = confounder_z_names
)

one_l_var <- as.character(pr_l$variable[1])
pr_l_one <- pr_l[pr_l$variable == one_l_var, ]
plot(pr_l_one$z, pr_l_one$est, type = "l",
     xlab = one_l_var, ylab = "Estimated h contribution",
     main = paste("Confounder model:", one_l_var))
lines(pr_l_one$z, pr_l_one$est + 1.96 * pr_l_one$se, lty = 2)
lines(pr_l_one$z, pr_l_one$est - 1.96 * pr_l_one$se, lty = 2)
```

To review all time-dependent confounder fits, loop over the nested list:

```{r all-confounder-fits, purl=FALSE}
for (tt in seq_along(fit$raw_results$fit_confounders)) {
  for (ll in seq_along(fit$raw_results$fit_confounders[[tt]])) {
    conf_fit <- first_bkmrfit(fit$raw_results$fit_confounders[[tt]][[ll]])
    z_names <- fit_z_names(conf_fit)
    print(bkmr::ExtractPIPs(conf_fit, sel = sel, z.names = z_names))
    bkmr::TracePlot(conf_fit, par = "beta", sel = sel,
                    main = paste("confounder t", tt, "model", ll))
  }
}
```

For binary outcomes or binary time-dependent confounders, these BKMR
summary plots describe the fitted probit/link-scale `h()` function. The
reported causal effect from `gbkmr_run()` is still returned on the
probability scale for binary outcomes.

## Causal Effect Draws

The g-computation stage stores posterior draws for the low and high
counterfactual means.

```{r ace-draws, purl=FALSE}
raw <- fit$raw_results
ace_draws <- raw$Yastar - raw$Ya

plot(ace_draws, type = "l",
     xlab = "Posterior draw",
     ylab = "ACE draw")

hist(ace_draws, breaks = 30,
     main = "Posterior ACE",
     xlab = "Y(a*) - Y(a)")

fit$causal_effect
```

For binary outcomes, these draws are on the risk-difference scale.

## Common Problems

| Symptom | Likely cause | Fix |
|---------|--------------|-----|
| Mixture columns are missing | `Z` has the wrong number of columns | Use `mixture_components * time_points` columns |
| Baseline covariates detected incorrectly | `X` columns are in the wrong order | Put all time-varying columns before baseline columns |
| Binary outcome uses continuous model | `outcome_type` left at the default | Set `outcome_type = "binary"` |
| fastBKMR requested for a binary outcome or binary time-varying confounder | Unsupported by the Gaussian-only fastBKMR path | Use `engine = "bkmr"` |
| Warnings about low effective sample size | Too few stable posterior draws | Increase `iter` and move `sel` later |
| Counterfactual values look implausible | Contrast outside support | Use `a_probs`, or choose `a_vals`/`astar_vals` within observed ranges |
| Parallel fastBKMR fails | Worker startup failed | Set `n_cores = 1` or run on a compute node |

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Large samples with fastBKMR](gbkmr_example_fastbkmr.html)
