---
title: "Diagnostics and Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette covers diagnostics for the current `causalBKMR` workflow.
The relevant user-facing functions are `prepare_gbkmr_data()`,
`detect_variable_patterns()`, and `gbkmr_run()`. Archived code under
`R/old/` is not part of this workflow.

## Check the Prepared Data

Before running MCMC, inspect the columns that `gbkmr_run()` will use.

```{r check-patterns}
patterns <- detect_variable_patterns(prepared, T = 3)
patterns

mixture_cols <- grep("^logM\\d+_\\d+$", names(prepared), value = TRUE)
mixture_cols
```

Confirm that:

- `patterns$p` is the expected number of mixture components per time.
- `mixture_cols` contains the expected `logM*_t` columns.
- `patterns$Ldim` is the expected number of time-varying covariates per time.
- `patterns$td_covariate_names` contains the expected covariate names.
- `patterns$td_vars_by_time` maps the follow-up covariate columns correctly.
- `Y` is present and points to the intended outcome column.

If the detected columns are wrong, return to the `Y`, `Z`, and `X`
matrices passed into `prepare_gbkmr_data()`. The most common issue is an
incorrect ordering of `X`: all time-varying covariates must come before
baseline covariates.

## Use the Input Audit

`gbkmr_run(verbose = TRUE)` prints an audit before fitting models.

```{r audit}
fit <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  engine = "auto",
  iter = 15000,
  K = 1000,
  verbose = TRUE
)
```

Read this audit before starting a long run. Check the outcome type,
number of time points, number of mixture components, selected engine,
sample size, and intervention contrast.

## Inspect Stored Diagnostics

The result object stores convergence diagnostics when available.

```{r stored-diagnostics}
fit$diagnostics
```

If warnings appear, rerun with more iterations and a later posterior
selection window:

```{r rerun-longer}
fit_long <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  iter = 60000,
  sel = seq(floor(60000 * 0.8), 60000, by = 25),
  K = 1000
)
```

The `sel` argument must contain MCMC iteration indices no larger than
`iter`. Use later iterations when early burn-in appears unstable.

## Inspect Raw Fits

`gbkmr_run()` returns lower-level model objects in `raw_results`.

```{r raw-results}
raw <- fit$raw_results

names(raw)
raw$fit_mediators
raw$fit_y
raw$meta
```

For the standard BKMR engine, `raw$fit_y` is a `bkmrfit` object. For
fastBKMR, `raw$fit_y` is a list of subset fits.

## Trace Plots

Use `bkmr::TracePlot()` for standard BKMR fits.

```{r trace-standard}
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "beta")
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "sigsq.eps")
bkmr::TracePlot(fit = fit$raw_results$fit_y, par = "r")
```

For fastBKMR, inspect subset fits:

```{r trace-fast}
fit_list <- fit$raw_results$fit_y

bkmr::TracePlot(fit = fit_list[[1]], par = "beta")

for (i in seq_along(fit_list)) {
  bkmr::TracePlot(fit = fit_list[[i]], par = "beta")
}
```

Long trends, abrupt jumps, or chains that remain stuck for long periods
are signs that the run needs more iterations or a different `sel`
window.

## Causal Effect Draws

The g-computation stage stores posterior draws for the low and high
counterfactual means.

```{r ace-draws}
raw <- fit$raw_results
ace_draws <- raw$Yastar - raw$Ya

plot(ace_draws, type = "l",
     xlab = "Posterior draw",
     ylab = "ACE draw")

hist(ace_draws, breaks = 30,
     main = "Posterior ACE",
     xlab = "Y(a*) - Y(a)")

fit$causal_effect
```

For binary outcomes, these draws are on the risk-difference scale.

## Common Problems

| Symptom | Likely cause | Fix |
|---------|--------------|-----|
| Mixture columns are missing | `Z` has the wrong number of columns | Use `mixture_components * time_points` columns |
| Baseline covariates detected incorrectly | `X` columns are in the wrong order | Put all time-varying columns before baseline columns |
| Binary outcome uses continuous model | `outcome_type` left at the default | Set `outcome_type = "binary"` |
| fastBKMR requested for a binary outcome | Unsupported engine/outcome combination | Use `engine = "bkmr"` |
| Warnings about low effective sample size | Too few stable posterior draws | Increase `iter` and move `sel` later |
| Counterfactual values look implausible | Contrast outside support | Use `a_probs`, or choose `a_vals`/`astar_vals` within observed ranges |
| Parallel fastBKMR fails | Worker startup failed | Set `n_cores = 1` or run on a compute node |

## See Also

- [Quick Start](gbkmr_quickstart.html)
- [Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Large samples with fastBKMR](gbkmr_example_fastbkmr.html)
